Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-10-16 paper.tex
Page break
Replace remaining mentions of foldable

  In defiance of Hinchliffe's rule, this article sets out to
  demonstrate that its title can be answered by the word ``yes''. We
  show that modern symmetric ciphers can be modeled through a small
  set of algebraic structures (Boolean algebra, Naperian and foldable
  functors). We reveal that these ciphers exhibit some interesting
  compositional structure at the type-level. This enables systematic
  code transformation, known in the cryptographic folklore as
  ``bitslicing'' and ``fixslicing''. Our work rests on a Coq
  development providing the specification of two ciphers, \Skinny{} and
  \Gift{}, deriving a bitsliced and fixsliced implementation for each and
  proving their correctness with respect to their specifications.

@@

The definitions of \coqe!mixcolumn_mod! is somewhat disappointing: we
would be hard-pressed to make any precise claim about the potentiality
of an efficient implementation. The current form is symbolically
useful, as it allows us to easily prove the correctness of fixslicing.
However, it remains a mathematical specification, not a piece of
software. In effect, we are now looking for a closed expression in the
language of Boolean algebras (\coqe!xor!, \etc{}), Naperian operations
(\coqe!init!, \coqe!lookup!) and foldable operations (more precisely,
\coqe!ror! and \coqe!rol!).

One solution\footnote{Needless to say, in reality, we first went
looking for 4 direct, simplified implementations, inspired
by~\citet{adominicai:fixslicing-gift}. Then we showed that these
implementations were equivalent to their respective
specification. Only later did we use Coq to understand how these
solutions \emph{could} have been derived from first principles.},
which consists in using Coq as a term rewriting engine, is to
postulate the existence of such an explicit form, let us call it
%
\coqe!mixcolumns_mod_explicit!.
%
We then claim that this definition ought to be equivalent to its
specification, \coqe!mixcolumns_mod! for $n \in \{0, 1, 2, 3\}$.
Making sure that the operations of Boolean algebra, Naperian and
foldable functors are opaque, $\beta$ reduction yields a term in our
target language. However, it suffers from significant redundancy: by
working over the term algebra, we have not quotiented out the
equational theory. To do so, we orient the equational theory in a
rewrite database and apply the \coqe!autosubst! tactics to obtain
simplified forms. Similarly, we retrieve convenient \coqe!let! forms
thanks to the \coqe!set (ident := term)! tactics. We are then left
with reading off the definition of \coqe!mixcolumns_mod_explicit! from
the proof goal.

@@

>>>>

  In defiance of Hinchliffe's rule, this article sets out to
  demonstrate that its title can be answered by the word ``yes''. We
  show that modern symmetric ciphers can be modeled through a small
  set of algebraic structures (Boolean algebra, Naperian and circulant
  functors). We reveal that these ciphers exhibit some interesting
  compositional structure at the type-level. This enables systematic
  code transformation, known in the cryptographic folklore as
  ``bitslicing'' and ``fixslicing''. Our work rests on a Coq
  development providing the specification of two ciphers, \Skinny{} and
  \Gift{}, deriving a bitsliced and fixsliced implementation for each and
  proving their correctness with respect to their specifications.

@@

The definitions of \coqe!mixcolumn_mod! is somewhat disappointing: we
would be hard-pressed to make any precise claim about the potentiality
of an efficient implementation. The current form is symbolically
useful, as it allows us to easily prove the correctness of fixslicing.
However, it remains a mathematical specification, not a piece of
software. In effect, we are now looking for a closed expression in the
language of Boolean algebras (\coqe!xor!, \etc{}), Naperian operations
(\coqe!init!, \coqe!lookup!) and circulant operations (more precisely,
\coqe!ror! and \coqe!rol!).

One solution\footnote{Needless to say, in reality, we first went
looking for 4 direct, simplified implementations, inspired
by~\citet{adominicai:fixslicing-gift}. Then we showed that these
implementations were equivalent to their respective
specification. Only later did we use Coq to understand how these
solutions \emph{could} have been derived from first principles.},
which consists in using Coq as a term rewriting engine, is to
postulate the existence of such an explicit form, let us call it
%
\coqe!mixcolumns_mod_explicit!.
%
We then claim that this definition ought to be equivalent to its
specification, \coqe!mixcolumns_mod! for $n \in \{0, 1, 2, 3\}$.
Making sure that the operations of Boolean algebra, Naperian and
circulant functors are opaque, $\beta$ reduction yields a term in our
target language. However, it suffers from significant redundancy: by
working over the term algebra, we have not quotiented out the
equational theory. To do so, we orient the equational theory in a
rewrite database and apply the \coqe!autosubst! tactics to obtain
simplified forms. Similarly, we retrieve convenient \coqe!let! forms
thanks to the \coqe!set (ident := term)! tactics. We are then left
with reading off the definition of \coqe!mixcolumns_mod_explicit! from
the proof goal.

@@

\clearpage
